Co-methylation review

### 2015-05-05 Rmarkdown/appendix.Rmd
Knocking off TODOs

\section{Chapter \ref{chap:co-methylation_review}}\label{app:co-methylation_review}

\subsection{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}

@@

\subsection{Estimating the number of CpGs used by \citet{Lacey:2013iy}}\label{app:lacey}

I want to estimate the number of CpGs used by \citet{Lacey:2013iy} to estimate the correlation of methylation levels. They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, _MTCTL2_. Unfortunately, there is no URL provided to these data. \citet{Lacey:2013iy} cite their data as having come from \citet{Tsumagari:2013gn}, which includes a sample "Normal Human Skeletal Muscle Myoblasts", _HSMM_ ([http://genome-mirror.duhs.duke.edu/ENCODE/cellTypes.html](http://genome-mirror.duhs.duke.edu/ENCODE/cellTypes.html), [http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/](http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/)). I believe this to be the same cell line. In any case, for my purposes I simply needed some comparable RRBS data, not necessarily the exact same sample as  _MTCTL2_.

There are approximately $60,000$ CpGs on chromosome 11 in these comparable RRBS data. The following R code shows how I computed this using three RRBS replicates of the _HSMM_ sample.

Firstly, I download the data and unzip it:
```{r, cache = FALSE, message = FALSE}
library(R.utils)
library(data.table)
dl_dir <- "../data/appendix/"
url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
files <- paste0('wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep',
                1:3, '.bed.gz')
dests <- paste0(dl_dir, files)
x <- mapply(function(f, d) {
  if (!file.exists(gsub(".gz", "", paste0(dl_dir, f)))) {
    download.file(url = paste0(url, f), d)
    gunzip(filename = d, overwrite = TRUE)
    }
  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t',
        verbose = FALSE)
  }, files, dests, SIMPLIFY = FALSE)
```

Then I count the number of CpGs on chromosome 11 in each replicate:
```{r}
y <- rbindlist(x)
y[V1 == "chr11", .N, by = list(V4)]
```

>>>>

\section{The probability that two dependent Bernoulli random variables are identical}\label{app:bernoulli_equal}

@@

### 2015-05-05 Rmarkdown/comethylation_review.Rmd
Fix table formatting
Knocking off TODOs

[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable RRBS libraries (see Appendix \ref{app:lacey}).

@@

\begin{table}[h]
\centering
\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple. For example, $n_{um}$ is the number of reads methylated at the first CpG and unmethylated at the second CpG. We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{u+} = n_{uu} + n_{um}$.}
\label{tab:landan}
\begin{tabular}{@{}cccc@{}}
\cmidrule(l){2-4}
 & \multicolumn{3}{c}{Second CpG} \\
 \cmidrule(l){2-4}
\multirow{3}{*}{First CpG} & & & \\
 & $n_{uu}$       & $n_{um}$       & $n_{u+}$ \\
 & $n_{mu}$       & $n_{mm}$       & $n_{m+}$ \\ \cmidrule(l){2-4}
 & $n_{+u}$  & $\cdot n_{+m}$  & $n_{++} = n$      \\
\end{tabular}
\end{table}

@@

A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$. Then, compute $R = cor(\bm{\beta})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci. This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data. As far as I am aware, `coMET` is the only software specifically designed to estimate between-sample co-methylation and integrate this with other genomic annotations.

>>>>

[^ncpgs]: The data for MTCTL2 are not publicly available so I estimated the number of sequenced CpGs on chromosome 11 from three comparable myoblast RRBS libraries (samples \texttt{wgEncode\-Haib\-Methyl\-Rrbs\-Hsmmtubefshd\-DukeSitesRep1}, \texttt{wgEncode\-Haib\-Methyl\-Rrbs\-Hsmmtubefshd\-DukeSitesRep2}, and \texttt{wgEncode\-Haib\-Methyl\-Rrbs\-Hsmmtubefshd\-DukeSitesRep3} available from \url{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/}).

@@

\begin{table}[!htb]
\centering
\caption[Counts of reads at a CpG 2-tuple]{Summary of read counts at a CpG 2-tuple. For example, $n_{um}$ is the number of reads methylated at the first CpG and unmethylated at the second CpG. We use the 'plus' notation to denote the sum over the index it replaces, e.g., $n_{u+} = n_{uu} + n_{um}$.}
\label{tab:landan}
\begin{tabular}{@{}ccccl@{}}
\cmidrule(l){2-5}
                           &              & \multicolumn{2}{c}{Second CpG} &              \\ \cmidrule(lr){3-4}
                           &              & Unmethylated    & Methylated   & Total        \\ \cmidrule(l){2-5}
\multirow{2}{*}{First CpG} & Unmethylated & $n_{uu}$        & $n_{um}$     & $n_{u+}$     \\
                           & Methylated   & $n_{mu}$        & $n_{mm}$   & $n_{m+}$     \\
                           & Total        & $n_{+u}$        & $n_{+m}$     & $n = n_{++}$ \\ \cmidrule(l){2-5}
\end{tabular}
\end{table}

@@

A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$. Then, compute $R = cor(\bm{\beta})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci. This is what the R package, `coMET` \citep{Martin:2015bs}, does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data. As far as I am aware, `coMET` is the only software specifically designed to estimate between-sample co-methylation and integrate this with other genomic annotations.
